---
title: "Data 607-Tidying and Transforming Data"
output: html_document
---


### Loading Libraries:

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)

```
### Introduction:

There are three datasets in this file:
1- Census data
2- Economic Research Service -U.S.Department of Agriculture ERS-USDA data
3- FIPS Code data
In this assignment, my aim is to perform data cleaning on all three data set, 
after initial analysis,merge thre three datasets on FIPS Code column and then 
find the co-relation between the variables and perform final analysis


# Dataset1:
### Reading census data from GitHub

```{r}
library(readr)

dataset1 <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_607_Project2/main/census_us_county_data.csv")

```


### Cleaning/Transforming census_us_county_data, dataset1:
```{r}
head(dataset1)
```

### Make the column names lowercase

```{r}
names(dataset1) <- tolower(names(dataset1))    # Convert colnames to lower case
dataset1  
```
### Seperating a column into two columns dataset1

```{r}
dataset1 <- dataset1 %>% separate(name, c("county", "state"), ",")

dataset1
```

### Dropping duplicates
```{r}
dataset1 <- dataset1 %>% distinct()

dataset1
```

### Renaming the columns
```{r}
names(dataset1)[4:9] <- c('median_age', 'household_income', 'per_capita_income', 'poverty_count',
                     'poverty_rate', 'unemployment_rate')

dataset1
```

### Removing the string 'county'
```{r}
dataset1$county <-gsub(" County","",as.character(dataset1$county))

dataset1
```

```{r}
glimpse(dataset1)
```

```{r}
str(dataset1)
```

```{r}
dataset1$state <- trimws(dataset1$state, which = c("left"))

str(dataset1)

```
### Sorting by columns
```{r}
# sort by state and county
# sort the dataframe in R using arrange

dataset1 <- arrange(dataset1,state, county)

dataset1
```
### fill NA with 0 in dataset1
```{r}
dataset1[is.na(dataset1)] <- 0

dataset1
```

### Analysis:
Step 1: Select data frame
Step 2: Group data
Step 3: Summarize the data
Step 4: Plot the summary statistics
```{r}
library(ggplot2)
# Step 1
dataset1 %>% 
#Step 2
group_by(state) 


```
### Multiple columns data by bar plot:
```{r}
# First filter the state Alabama and selecting the columns 

dataset1_alabama <- dataset1 %>% filter(state == "Alabama") %>%

  select(c(3:9, 1))

dataset1_alabama
```
```{r}
str(dataset1_alabama)
```


### Transform the table from wide form to long form
```{r}
data1_al <- dataset1_alabama %>%

  pivot_longer(cols = c(2:8),

               names_to = "Variable",

               values_to = "Value")

data1_al
```

```{r}
bar_plot <- ggplot(data = data1_al, aes(x = Variable, y = Value, fill = Variable)) +

  geom_bar(stat = "identity") + coord_flip() +
  
  labs(
        x = "Variables",
        y = "Values",
        title = paste(
            "Census data for Alabama"
        )
)
bar_plot

```

# Dataset 2:

### Reading the dataset 2 from GitHub
```{r}
dataset2 <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_607_Project2/main/ers_usda_education_data.csv")
```


### Cleaning/Transforming ers_usda_education_data, dataset2:
```{r}
head(dataset2)
```
```{r}
colnames(dataset2)
```


### Dropping 4:32 columns
```{r}

dataset2 <- select(dataset2, -(4:32))
                           

dataset2
```
### Rename the columns by removing space with underscore
```{r}
names(dataset2) <- gsub(" ", "_", names(dataset2))

dataset2
```
### Removing  comma from column names
```{r}
names(dataset2) <- gsub(",", "", names(dataset2))

dataset2
```
### Removing  apostrophy from column names
```{r}
names(dataset2) <- gsub("'", "", names(dataset2))

dataset2
```
### Replacing dash with underscore in column names

```{r}
names(dataset2) <- gsub("-", "_", names(dataset2))

dataset2
```


### Making column names to lowercase
```{r}
names(dataset2) <- tolower(names(dataset2))    # Convert colnames to lower case
dataset2
```

### Removing the string  'county' from 'area_name'
```{r}
dataset2$area_name <-gsub(" County","",as.character(dataset2$area_name))

dataset2
```
### Dropping the duplicate rows
```{r}
dataset2 <- dataset2 %>% distinct()

dataset2
```
### Drop first row
```{r}
dataset2 <- dataset2[-c(seq(1, 1, by=0)),]

dataset2
```
### fill NA with 0 in dataset3
```{r}
dataset2[is.na(dataset2)] <- 0

dataset2
```

```{r}
names(dataset2)[3] <- c('county')

dataset2
```

```{r}
# sort by state and county
# sort the dataframe in R using arrange

dataset2 <- arrange(dataset2,state, county)

dataset2
```
```{r}
glimpse(dataset2)
```



# Dataset 3:

### Reading the data from Github:

```{r}
dataset3 <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_607_Project2/main/county_fips_data.csv")
```


### Tidying/Transforming dataset3
```{r}
# Now tidying dataset3

head(dataset3)
```

### Making column names lowercase
```{r}
names(dataset3) <- tolower(names(dataset3))    # Convert colnames to lower case
dataset3
```

### Removing the string 'county' from 'county_name' column values

```{r}
dataset3$county_name <-gsub(" County","",as.character(dataset3$county_name))

dataset3
```
### Rename county_name to county

```{r}
dataset3 <- rename(dataset3, county = county_name)

dataset3
```


### Fill NA with 0
```{r}
dataset3[is.na(dataset3)] <- 0
```


```{r}
str(dataset3)
```

```{r}
glimpse(dataset3)
```

### Performing left join between Census and FIPS_Code data on 'County' data:

```{r}
joint_df <- dataset2 %>% inner_join(dataset3,by="county")

joint_df
```
### Now joining Census dataset1 to jointdataset

```{r}
#combined_df<- merge(dataset1, joint_df, by = 'county', all.x= TRUE)
combined_df = dataset1 %>% inner_join(joint_df,by="county")

combined_df
```


### Use the 'select' and 'one_of' functions of the dplyr package to delete columns

```{r}
col_remove <- c("state.y", "fips")     # Define columns that should be dropped

```

### Removing the repeated column
```{r}

combined_df <- combined_df %>%            # Apply select & one_of functions
  select(- one_of(col_remove))

combined_df
```

### Rename state.x column to state
 
```{r}
combined_df <- rename(combined_df, state = state.x)

combined_df
```

### Remove duplicate rows of the dataframe

```{r}
combined_df <- combined_df[!duplicated(combined_df[,c('county')]),]

combined_df
```


# Statistical Analysis:

```{r}
summary(combined_df)
```
### Find the maximun percapita income state

```{r}
# Get Maximum value of the column by column name
 
max(combined_df$per_capita_income)
```

### Move column state to 1st position

```{r}
combined_df <- combined_df %>% relocate(state, .before = population)

combined_df
```

### Group by state and finfing the average of the following indicators

```{r}
by_state <- combined_df %>% group_by(state)

by_state <- by_state %>% summarise(
  population = mean(population),
  per_capita_income = mean(per_capita_income),
  poverty_count = mean(poverty_count),
  bachelors_degree = mean(bachelors_degree_or_higher_2015_19)
)

by_state
```

### To find the highest poverty_count by state

```{r}
library(ggplot2)

ggplot(by_state, aes(fill=state, y=poverty_count, x=state)) +
  geom_bar(position='dodge', stat='identity',
           color = "black", width = 0.60)


```
### To find which state exactly has the highest poverty_count in the above barplot since the sate is not detectable easily
It shows District of Columbia has the highest poverty count
```{r}
by_state %>% slice(which.max(poverty_count))
```

### Now repeat the above graph for education

```{r}
library(ggplot2)

ggplot(by_state, aes(fill=state, y=bachelors_degree, x=state)) +
  geom_bar(position='dodge', stat='identity',
           color = "black", width = 0.60)


```
### To find which state exactly has the highest education average in the above barplot since the sate is not detectable easily
It shows District of Columbia also has the highest bachelors_degree average
```{r}
by_state %>% slice(which.max(bachelors_degree))
```
### We can clearly see that there is a maximum increase for poverty and education for the state of District of Columbia
```{r}
# We can that 

by_state %>%
  ggplot(aes(x=poverty_count, 
             y=bachelors_degree, 
             color=state))+
  geom_point()+
  geom_smooth(method="lm",se = FALSE)
ggsave("add_regression_line_per_state_to_scatterplot_ggplot2.png")

```






